---
title: "Defining a penalty"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Defining a penalty}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>")
library(penalties7)
set.seed(1)
```

A penalty is $\rho(D\beta;\theta)$: a linear map, a scalar function of the
mapped coefficients, and hyperparameters of its own. Almost every R package
that fits a penalized model writes one inside the function that fits it, so
the penalty is a term in an objective rather than an object, and nothing can
ask it whether it has a kink, what its rank is, or how its value moves with
its hyperparameter. Those are the questions a fitting layer has to answer
before it can choose a scheme.

This vignette builds a penalty the package does not ship and shows what it has
to supply and what it gets in return. The example is the Berhu, the reverse
Huber of Owen (2007):

$$\rho(\beta) = \begin{cases}
  \lambda\lvert\beta\rvert, & \lvert\beta\rvert \le \delta,\\[2pt]
  \lambda\dfrac{\beta^2 + \delta^2}{2\delta}, & \lvert\beta\rvert > \delta,
\end{cases}$$

the lasso near the origin and the ridge away from it. It has a kink at zero, so
it selects; it grows quadratically in the tails, so a large coefficient is not
shrunk proportionally.

## The class and its constructor

```{r}
BerhuPenalty <- S7::new_class("BerhuPenalty", parent = penalty)

berhu_penalty <- function(n_coef, map = NULL) {
  BerhuPenalty(
    penalty_name = "berhu",
    map = map,
    n_coef = as.integer(n_coef),
    params = c("lambda", "delta"),
    params_bounds = list(lambda = c(0, Inf), delta = c(0, Inf)),
    link_params = list(lambda = linkfunctions7::log_link(),
                       delta = linkfunctions7::log_link()),
    params_smooth = c(lambda = TRUE, delta = TRUE)
  )
}
```

`params_bounds` are open intervals and `link_params` says how each
hyperparameter reaches the unconstrained scale an outer search steps on.
`params_smooth` records which of them the value is differentiable in, so a
numerical guard can switch itself off where it would be measuring a kink.

## The methods

Every generic aligns `theta` before it dispatches, reordering it to
`pen@params`, stripping stray names off the values and checking it against
`params_bounds`. A method therefore receives a list already in the family's own
order and does not align it again. One helper reads it and works out the branch
each coefficient falls in, so the eight methods share it:

```{r}
.berhu <- function(pen, beta, theta) {
  list(lam = theta$lambda, del = theta$delta,
       a = abs(beta), big = abs(beta) > theta$delta)
}
```

The value, the gradient in $\beta$ and the Hessian in $\beta$:

```{r}
S7::method(penalty_value, BerhuPenalty) <- function(pen, beta, theta, ...) {
  p <- .berhu(pen, beta, theta)
  sum(ifelse(p$big, p$lam * (beta^2 + p$del^2) / (2 * p$del), p$lam * p$a))
}

S7::method(penalty_gradient, BerhuPenalty) <- function(pen, beta, theta, ...) {
  p <- .berhu(pen, beta, theta)
  ifelse(p$big, p$lam * beta / p$del, p$lam * sign(beta))
}

S7::method(penalty_hessian, BerhuPenalty) <- function(pen, beta, theta, ...) {
  p <- .berhu(pen, beta, theta)
  diag(ifelse(p$big, p$lam / p$del, 0), length(beta))
}
```

Then the three blocks in the hyperparameters, which are what a marginal
criterion needs in order to estimate them. Note the `scale` argument: an S7
method's formals have to include the generic's named arguments, and these three
generics carry one.

```{r}
S7::method(penalty_grad_theta, BerhuPenalty) <- function(
    pen, beta, theta, scale = c("parameter", "link"), ...) {
  p <- .berhu(pen, beta, theta)
  list(lambda = sum(ifelse(p$big, (beta^2 + p$del^2) / (2 * p$del), p$a)),
       delta  = sum(ifelse(p$big,
                           p$lam * (p$del^2 - beta^2) / (2 * p$del^2), 0)))
}

S7::method(penalty_hess_theta, BerhuPenalty) <- function(
    pen, beta, theta, scale = c("parameter", "link"), ...) {
  p <- .berhu(pen, beta, theta)
  list(lambda_lambda = 0,
       delta_delta   = sum(ifelse(p$big, p$lam * beta^2 / p$del^3, 0)),
       lambda_delta  = sum(ifelse(p$big,
                                  (p$del^2 - beta^2) / (2 * p$del^2), 0)))
}

S7::method(penalty_cross, BerhuPenalty) <- function(
    pen, beta, theta, scale = c("parameter", "link"), ...) {
  p <- .berhu(pen, beta, theta)
  list(lambda = ifelse(p$big, beta / p$del, sign(beta)),
       delta  = ifelse(p$big, -p$lam * beta / p$del^2, 0))
}
```

Finally the two structural questions. `penalty_kinks()` returns the **positions**
of the kinks, not their number; the Berhu's only kink is the origin, where SCAD
has five.

```{r}
S7::method(penalty_kinks, BerhuPenalty) <- function(pen, theta, ...) 0
S7::method(is_proper, BerhuPenalty) <- function(pen) TRUE

pen <- berhu_penalty(5)
b <- c(-2.5, -0.4, 0, 0.7, 3.1)
th <- list(lambda = 1.3, delta = 1)

penalty_value(pen, b, th)
penalty_gradient(pen, b, th)
diag(penalty_hessian(pen, b, th))
```

### The shapes are part of the contract

Every hyperparameter block is a **named list**, not a matrix or a data frame:
`penalty_grad_theta()` one scalar per hyperparameter, `penalty_hess_theta()`
keyed `"lambda_lambda"`, `"delta_delta"`, `"lambda_delta"`, and
`penalty_cross()` one vector of length `n_coef` per hyperparameter. Compare a
shipped penalty and the shapes are the same:

```{r}
str(penalty_hess_theta(scad_penalty(n_coef = 3), c(-2, 0.3, 1.5),
                       list(lambda = 1, a = 3.7)))
```

## Validating it

`check_penalty()` differentiates the value numerically and compares, one block
at a time:

```{r}
print(check_penalty(pen, theta = th, verbose = FALSE))
```

The reference is `numDeriv`, which shares no arithmetic with the formulas, and
the cross block is checked by Richardson extrapolation on the analytic gradient,
never by differencing the value twice.

### A numerical reference is invalid at a kink

Hand the validator a $\beta$ with a coordinate sitting exactly on the kink and
the second-derivative check fails on correct code:

```{r}
check_penalty(pen, beta = b, theta = th, verbose = FALSE)[1:2, ]
```

`b` contains an exact zero, the gradient jumps from $-\lambda$ to $+\lambda$
there, and a central difference across the jump measures the kink, not
the formula. When `beta` is left `NULL` the validator draws it and nudges each
draw off any position `penalty_kinks()` declares, which is why the run above
passed. Moving the zero is enough:

```{r}
check_penalty(pen, beta = c(-2.5, -0.4, 0.05, 0.7, 3.1),
              theta = th, verbose = FALSE)[1:2, ]
```

This is the reason `penalty_kinks()` returns positions. A count would say a kink
exists; the positions say where a reference must not be taken.

## What the base class supplies, and what it refuses

```{r}
c(is_quadratic   = is_quadratic(pen),
  beta_quadratic = beta_quadratic(pen),
  has_prox       = has_prox(pen))
try(penalty_rank(pen))
```

`penalty_rank()` and `penalty_matrix()` are refused because there is no matrix:
a rank is a property of a quadratic form, and returning one for a penalty that
is not quadratic would be a different object under the same name.

`has_prox()` is `FALSE` because no proximal operator has been registered, and
that is the answer with consequences. A fitting layer routes a block by whether
its penalty is twice differentiable: a smooth one joins the system solved by
`iwls()` or an ordinary optimizer, and a kinked one needs a proximal step or a
coordinate descent, both of which need the operator. A kinked penalty with no
`penalty_prox()` can be evaluated and differentiated away from its kinks, and
cannot be fitted by the scheme its kink calls for.

The Berhu's operator is closed, so registering one is a further exercise, not an
obstruction; `penalty_prox_spec()` is the piecewise-linear table a
compiled coordinate descent reads, and supplying it is what puts a penalty on
that path.

## The normalizing constant is kept

A separable penalty built from a distributions7 family is the negative
log-density of that family, and the shipped ones keep the constant. The lasso is
a fixed Laplace in its rate chart, so its value carries $-\log(\lambda/2)$ per
coefficient:

```{r}
small <- c(-0.3, 0.2)
c(berhu = penalty_value(pen, small, th),
  lasso = penalty_value(lasso_penalty(n_coef = 2), small,
                        list(lambda = 1.3)),
  constant = -2 * log(1.3 / 2))
```

Dropping the constant makes hyperparameter estimation degenerate: without it the
objective falls monotonically as $\lambda \to 0$, and a $t$ prior's degrees of
freedom cannot be estimated at all. The Berhu above has no constant because it
is not a normalized density; a penalty of that kind states its hyperparameters
as what they are, quantities the caller sets or an outer criterion chooses on a
prediction error instead of a marginal likelihood.

## Summary

* **Minimum to define a penalty:** a subclass of `penalty` and methods for
  `penalty_value()`, `penalty_gradient()`, `penalty_hessian()`,
  `penalty_grad_theta()`, `penalty_hess_theta()`, `penalty_cross()`,
  `penalty_kinks()` and `is_proper()`.
* The three hyperparameter blocks take a `scale` argument, because their
  generics declare one, and every one returns a **named list**.
* `penalty_kinks()` returns positions, so a numerical reference can be kept off
  them; a count would not be enough.
* `check_penalty()` compares against `numDeriv` and Richardson, and draws a
  $\beta$ nudged off every declared kink when it is given none.
* `is_quadratic()`, `penalty_rank()`, `penalty_matrix()`, `penalty_logpdet()`,
  `penalty_prox()` and `penalty_prox_spec()` come from the base class, and the
  first three refuse where the penalty is not quadratic.
* `has_prox()` decides which fitting scheme a kinked block can reach.
